ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-10-09 pyked.tex
extended RCM example, with figs

\usepackage{fancyvrb,newverbs,xcolor}

@@

In Python, additional functionality can be imported into a script or session by the \pybox|import|
keyword. Cantera, NumPy, and PyKED must be imported into the session so that we can work with the
code. In the case of Cantera and NumPy, we will use many functions from these libraries, so we
assign them abbreviations (\pybox|ct| and \pybox|np|, respectively) for convenience. From PyKED, we
will only be using the \pybox|ChemKED| class, so this is all that is imported:
%
\begin{pythonbox}
import cantera as ct
import numpy as np
from pyked import ChemKED
\end{pythonbox}
%
Next, we have to load the ChemKED file and retrieve the first element of the \pybox|datapoints|
list. Although this file only encodes a single experiment, the \pybox|datapoints| attribute will
always be a list (in this case, of length 1). As mentioned previously, the elements of the
\pybox|datapoints| list are instances of the \pybox|DataPoint| class, which we store in the variable
\pybox|dp|.
%
\begin{pythonbox}
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox|DataPoint| class. \pk{} uses instances of the Pint \pybox|Quantity| class to
store values with units, while Cantera expects a floating-point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units, and we use the \pybox|magnitude|
attribute of the \pybox|Quantity| class to take only the numerical part. We also retrieve the
initial mixture mole fractions in a format Cantera will understand:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()
\end{pythonbox}
%
With these properties defined, we have to create the objects in Cantera that represent the physical
state of the system to be studied. In Cantera, the \pybox|Solution| class stores the thermodynamic,
kinetic, and transport data from an input file in the CTI format. After the \pybox|Solution| object
is created, we can set the initial temperature, pressure, and mole fractions using the \pybox|TPX|
attribute of the \pybox|Solution| class:

\begin{pythonbox}
# Load the mechanism and set the initial state of the mixture
gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_initial, P_initial, X_initial
\end{pythonbox}

With the thermodynamic and kinetic data loaded and the initial conditions defined, we  need to
install the \pybox|Solution| instance into an \pybox|IdealGasReactor| which implements the equations
for mass, energy, and species conservation. In addition, we create a \pybox|Reservoir| to represent
the environment external to the reaction chamber. The input file used for the environment,
\verb|air.xml|, is included with Cantera and represents an average composition of air.

\begin{pythonbox}
# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox|IdealGasReactor|, a \pybox|Wall| must be
installed between the reactor and environment and assigned a velocity. The \pybox|Wall| allows the
environment to do work on the reactor (or vice versa) and change the reactor's thermodynamic state;
we use a \pybox|Reservoir| for the environment because in Cantera, \pybox|Reservoir|s always have a
constant thermodynamic state and composition. Using a \pybox|Reservoir| accelerates the solution
compared to using two \pybox|IdealGasReactor|s, since the composition and state of the environment
are typically not necessary for the solution of autoignition problems. Although we do not show the
details here, a reference implementation of a class that computes the wall velocity given the volume
history of the reactor is available in CanSen~\autocite{cansen}, in the
\pybox|cansen.profiles.VolumeProfile| class.
%
\begin{pythonbox}
# Retrieve the time and volume from the history in the datapoint
exp_time = dp.volume_history.time.magnitude
exp_volume = dp.volume_history.volume.magnitude
keywords = {'vproTime': exp_time, 'vproVol': exp_volume}
# Install the Wall between the reactor and environment
ct.Wall(reac, env, velocity=VolumeProfile(keywords))
\end{pythonbox}

Then, the \pybox|IdealGasReactor| is installed in a \pybox|ReactorNet|. The \pybox|ReactorNet|
implements the connection to the numerical solver (CVODES \autocite{hindmarsh_sundials:_2005} is
used in Cantera) to solve the energy and species equations. For this example, it is best practice
to set the maximum time step allowed in the solution to be the minimum time difference in the time array from the volume trace:

\begin{pythonbox}
netw = ct.ReactorNet([reac])
netw.set_max_time_step(np.min(np.diff(exp_time)))
\end{pythonbox}

To calculate the ignition delay, we will follow the definition specified in the \ck{} file for
this experiment, where the experimentalists used the maximum of the time derivative of the pressure
to define the ignition delay. To calculate this derivative, we need to store the state variables and the composition on each time step, so we initialize several Python lists to act as storage:

\begin{pythonbox}
# Initialize lists to store solution information
time = []
temperature = []
pressure = []
volume = []
mass_fractions = []
\end{pythonbox}

Finally, the problem is integrated using the \pybox|step| method of the \pybox|ReactorNet|. The
\pybox|step| method takes one timestep forward on each call, with step size determined by the CVODES
solver (CVODES uses an adaptive time-stepping algorithm). On each step, we add the relevant
variables to their respective lists. The problem is integrated until a user-specified end time, in
this case \SI{50}{\milli\second}, although in principle, the user could end the simulation on any
condition they choose:

\begin{pythonbox}
# Integrate for 50 ms
while netw.time < 0.05:
    time.append(netw.time)
    temperature.append(reac.T)
    pressure.append(reac.thermo.P)
    volume.append(reac.volume)
    mass_fractions.append(reac.Y)
    netw.step()
\end{pythonbox}

At this point, the user would post-process the information in the \pybox|pressure| list to calculate
the derivative by whatever algorithm they choose.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.7\textwidth]{ignition-delays.pdf}
    \caption{Comparison of experimental and simulated ignition delays of
    \textit{tert}-butanol from shock tube example.}
    \label{fig:ignition-delays}
\end{figure}

>>>>

\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage{textcomp}
\usepackage{latexsym,amsmath,amssymb}

\usepackage{fancyvrb,newverbs,xcolor}
\usepackage{graphicx}
\usepackage{caption,subcaption}

@@

In Python, additional functionality can be imported into a script or session by the \pybox|import|
keyword. Cantera, NumPy, and PyKED must be imported into the session so that we can work with the
code. In the case of Cantera and NumPy, we will use many functions from these libraries, so we
assign them abbreviations (\pybox|ct| and \pybox|np|, respectively) for convenience. From PyKED, we
will only be using the \pybox|ChemKED| class, so this is all that is imported:
%
\begin{pythonbox}
import cantera as ct
import numpy as np
from pyked import ChemKED
\end{pythonbox}
%
Next, we have to load the ChemKED file and retrieve the first element of the \pybox|datapoints|
list. Although this file only encodes a single experiment, the \pybox|datapoints| attribute will
always be a list (in this case, of length 1). As mentioned previously, the elements of the
\pybox|datapoints| list are instances of the \pybox|DataPoint| class, which we store in the variable
\pybox|dp|.
%
\begin{pythonbox}
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox|DataPoint| class. \pk{} uses instances of the Pint \pybox|Quantity| class to
store values with units, while Cantera expects a floating-point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units, and we use the \pybox|magnitude|
attribute of the \pybox|Quantity| class to take only the numerical part. We also retrieve the
initial mixture mole fractions in a format Cantera will understand:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()
\end{pythonbox}

With these properties defined, we have to create the objects in Cantera that represent the physical
state of the system to be studied. In Cantera, the \pybox|Solution| class stores the thermodynamic,
kinetic, and transport data from the GRI Mech 3.0~\cite{GRI3.0} model included with Cantera.
After the \pybox|Solution| object is created, we can set the initial temperature, pressure,
and mole fractions using the \pybox|TPX| attribute of the \pybox|Solution| class:
%
\begin{pythonbox}
# Load the mechanism and set the initial state of the mixture
gas = ct.Solution('gri30.cti')
gas.TPX = T_initial, P_initial, X_initial
\end{pythonbox}

With the thermodynamic and kinetic data loaded and the initial conditions defined, we  need to
install the \pybox|Solution| instance into an \pybox|IdealGasReactor| which implements the equations
for mass, energy, and species conservation. In addition, we create a \pybox|Reservoir| to represent
the environment external to the reaction chamber. The input file used for the environment,
\verb|air.xml|, is included with Cantera and represents an average composition of air.
%
\begin{pythonbox}
# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox|IdealGasReactor|, a \pybox|Wall| must be
installed between the reactor and environment and assigned a velocity. The \pybox|Wall| allows the
environment to do work on the reactor (or vice versa) and change the reactor's thermodynamic state;
we use a \pybox|Reservoir| for the environment because in Cantera, \pybox|Reservoir|s always have a
constant thermodynamic state and composition. Using a \pybox|Reservoir| accelerates the solution
compared to using two \pybox|IdealGasReactor|s, since the composition and state of the environment
are typically not necessary for the solution of autoignition problems. Although we do not show the
details here, a reference implementation of a class that computes the wall velocity given the volume
history of the reactor is available in CanSen~\autocite{cansen}, in the
\pybox|cansen.profiles.VolumeProfile| class.
%
\begin{pythonbox}
from cansen.profiles import VolumeProfile
# Retrieve the time and volume from the history in the datapoint
exp_time = dp.volume_history.time.magnitude
exp_volume = dp.volume_history.volume.magnitude
keywords = {'vproTime': exp_time, 'vproVol': exp_volume}
# Install the Wall between the reactor and environment
ct.Wall(reac, env, velocity=VolumeProfile(keywords))
\end{pythonbox}

Then, the \pybox|IdealGasReactor| is installed in a \pybox|ReactorNet|. The \pybox|ReactorNet|
implements the connection to the numerical solver (CVODES \autocite{hindmarsh_sundials:_2005} is
used in Cantera) to solve the energy and species equations. For this example, it is best practice
to set the maximum time step allowed in the solution to be the minimum time difference in
the time array from the volume trace:
%
\begin{pythonbox}
netw = ct.ReactorNet([reac])
netw.set_max_time_step(np.min(np.diff(exp_time)))
\end{pythonbox}

To calculate the ignition delay, we will follow the definition specified in the \ck{} file for
this experiment, where the experimentalists used the maximum of the time derivative of the pressure
to define the ignition delay. To calculate this derivative, we need to store the state variables
and the composition on each time step, so we initialize several Python lists to act as storage:
%
\begin{pythonbox}
# Initialize lists to store solution information
time = []
temperature = []
pressure = []
volume = []
mass_fractions = []
\end{pythonbox}

Finally, the problem is integrated using the \pybox|step| method of the \pybox|ReactorNet|. The
\pybox|step| method takes one timestep forward on each call, with step size determined by the CVODES
solver (CVODES uses an adaptive time-stepping algorithm). On each step, we add the relevant
variables to their respective lists. The problem is integrated until a user-specified end time, in
this case \SI{50}{\milli\second}, although in principle, the user could end the simulation on any
condition they choose:
%
\begin{pythonbox}
# Integrate for 50 ms
while netw.time < 0.05:
    time.append(netw.time)
    temperature.append(reac.T)
    pressure.append(reac.thermo.P)
    volume.append(reac.volume)
    mass_fractions.append(reac.Y)
    netw.step()
\end{pythonbox}

At this point, the user would post-process the information in the \pybox|pressure| list to calculate
the derivative by whatever algorithm they choose. Here, we plot pressure versus time from the
simulation, and compare the simulated and experimental volume traces,
using the Matplotlib library~\cite{matplotlib} and shown in Figure~\ref{fig:rcm}:
%
\begin{pythonbox}
import matplotlib.pyplot as plt

plt.figure()
plt.plot(time, pressure)
plt.ylabel('Pressure [Pa]')
plt.xlabel('Time [s]')

plt.figure()
plt.plot(exp_time, exp_volume/exp_volume[0], label='Experimental volume', linestyle='--')
plt.plot(time, volume, label='Simulated volume')
plt.legend(loc='best')
plt.ylabel('Volume [m^3]')
plt.xlabel('Time [s]')
\end{pythonbox}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=\linewidth]{rcm-pressure.pdf}
        \caption{Simulated pressure trace}
    \end{subfigure}
    \\
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=\linewidth]{rcm-volume.pdf}
        \caption{Experimental and simulated volume histories}
    \end{subfigure}
    \caption{Results from RCM example}
    \label{fig:rcm}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.7\textwidth]{ignition-delays.pdf}
    \caption{Comparison of experimental and simulated ignition delays of
    \textit{tert}-butanol from shock tube example.}
    \label{fig:ignition-delays}
\end{figure}
